---
title: "ggplot Wizardry"
author: "Onkar Singh"
date: "3/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Tutorial](https://z3tt.github.io/OutlierConf2021/) [Slides](https://www.cedricscherer.com/slides/OutlierConf2021_ggplot-wizardry.pdf) [Codes](https://github.com/Z3tt/OutlierConf2021/blob/main/R/OutlierConf2021_ggplotWizardry_HandsOn.Rmd) [Recordings](https://www.youtube.com/watch?v=7UjA_5gNvdw&list=PLAm5TIX-yz7IkKOUcStM_vl8AD0S9v0co&index=34)

A Step-by-Step tutorial as supplement to my talk "ggplot Wizardry: My Favorite Tricks and Secrets for Beautiful Plot in R" at OutlierConf 2021.

## **Packages**

```{r}
library(tidyverse)   ## data science package collection (incl. the ggplot2 package)
library(systemfonts) ## use custom fonts (need to be installed on your OS)  
library(scico)       ## scico color palettes(http://www.fabiocrameri.ch/colourmaps.php) in R 
library(ggtext)      ## add improved text rendering to ggplot2
library(ggforce)     ## add missing functionality to ggplot2
library(ggdist)      ## add uncertainity visualizations to ggplot2
library(magick)      ## load images into R
library(patchwork)   ## combine outputs from ggplot2
```

## **Data**

Data were collected and made available by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER, a member of the Long Term Ecological Research Network.

The data is easily available as well as `palmerpenguins` package. The goal of `palmerpenguins` is to provide a great dataset for data exploration & visualization, as an alternative to `iris`.

→ [allisonhorst.github.io/palmerpenguins](https://allisonhorst.github.io/palmerpenguins)

```{r}
penguins <- 
  readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-28/penguins.csv') %>% 
  ## correct species name
  mutate(species = if_else(species == "Adelie", "Adélie", species)) %>% 
  ## remove missing observations
  filter(!is.na(bill_length_mm), !is.na(bill_depth_mm))

## if you like: nicely formatted table in report
penguins %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = TRUE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive")) %>%
  kableExtra::scroll_box(width = "700px", height = "500px")
```

# A Basic ggplot

```{r}
## simple plot: data + mappings + geometry
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(alpha = .6, size = 3.5)

ggsave("00_scatterplot_raw.pdf", width = 9, height = 8, device = cairo_pdf)
```

# A Customized ggplot

```{r}
## change global theme settings (for all following plots)
theme_set(theme_minimal(base_size = 12, base_family = "Open Sans"))

## modify plot elements globally (for all following plots)
theme_update(
  axis.ticks = element_line(color = "grey92"),
  axis.ticks.length = unit(.5, "lines"),
  panel.grid.minor = element_blank(),
  legend.title = element_text(size = 12),
  legend.text = element_text(color = "grey30"),
  plot.title = element_text(size = 18, face = "bold"),
  plot.subtitle = element_text(size = 12, color = "grey30"),
  plot.caption = element_text(size = 9, margin = margin(t = 15))
)
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = body_mass_g), alpha = .6, size = 3.5) + 
  ## custom axes scaling
  scale_x_continuous(breaks = 3:6 * 10, limits = c(30, 60)) +
  scale_y_continuous(breaks = seq(12.5, 22.5, by = 2.5), limits = c(12.5, 22.5)) +
  ## custom colors
  scico::scale_color_scico(palette = "bamako", direction = -1) +
  ## custom labels
  labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (Pygoscelis)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) PLoS ONE',
    x = 'Bill Length (mm)', 
    y = 'Bill Depth (mm)',
    color = 'Body mass (g)'
  )

ggsave("01_theme_color_labs.pdf", width = 9, height = 8, device = cairo_pdf)
```

# `{ggtext}`

> The `ggtext` package provides simple Markdown and HTML rendering for ggplot2. Under the hood, the package uses the `gridtext` package for the actual rendering, and consequently it is limited to the feature set provided by `gridtext`.\
> Support is provided for Markdown both in theme elements (plot titles, subtitles, captions, axis labels, legends, etc.) and in geoms (similar to `geom_text()`). In both cases, there are two alternatives, one for creating simple text labels and one for creating text boxes with word wrapping.

→ [wilkelab.org/ggtext](https://wilkelab.org/ggtext)

## **`element_markdown()`**

`element_markdown()` → formatted text elements, e.g. titles, caption, axis text, striptext

```{r}
## assign plot to `g` - we can ad new things to this plot later
## (wrapped in parenthesis so it is assigned and plotted in one step)
(gt <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = body_mass_g), alpha = .6, size = 3.5) + 
  scale_x_continuous(breaks = 3:6 * 10, limits = c(30, 60)) +
  scale_y_continuous(breaks = seq(12.5, 22.5, by = 2.5), limits = c(12.5, 22.5)) +
  scico::scale_color_scico(palette = "bamako", direction = -1) +
  ## markdown formatting using asterisks
  labs(
    title = 'Bill Dimensions of Brush-Tailed Penguins (*Pygoscelis*)',
    subtitle = 'A scatter plot of bill depth versus bill length.',
    caption = 'Data: Gorman, Williams & Fraser (2014) *PLoS ONE*',
    x = '**Bill Length** (mm)', 
    y = '**Bill Depth** (mm)',
    color = 'Body mass (g)'
  ) +
  ## render respective text elements
  theme(
    plot.title = ggtext::element_markdown(),
    plot.caption = ggtext::element_markdown(),
    axis.title.x = ggtext::element_markdown(),
    axis.title.y = ggtext::element_markdown()
  )
)
 
ggsave("02a_ggtext_element_markdown.pdf", width = 9, height = 8, device = cairo_pdf)
```

## **`element_markdown()` in combination with HTML**

```{r}
## use HTML syntax to change text color
gt_mar <- gt +
  labs(title = 'Bill Dimensions of Brush-Tailed Penguins <i style="color:#28A87D;">Pygoscelis</i>') +
  theme(plot.margin = margin(t = 25))

ggsave("02b_ggtext_element_markdown_color.pdf", width = 9, height = 8, device = cairo_pdf)

```

```{r}
## use HTML syntax to change font and text size
gt_mar +
  labs(title = 'Bill Dimensions of Brush-Tailed Penguins <b style="font-size:32pt;font-family:blacksword;">Pygoscelis</b>')

ggsave("02c_ggtext_element_markdown_font.pdf", width = 9, height = 8, device = cairo_pdf)
```
